A Methodology and Implementation of Automated Emissions Harmonization for Use in Integrated Assessment Models

### 2017-06-06 future.tex
some final edits

However, this work provides a new direction and framework in which the IAM and
climate communities can follow in order to reduce the necessity of expert
opinion and increase transparency and reproducibility of harmonization
exercises. Furthermore, it provides an open-source, tested, and documented code
base which can be used and improved upon by these communities. Both of these are
clear steps in a positive direction for future climate and integrated assessment
modeling exercises.

>>>>

However, this work provides a new direction and framework which the IAM and
climate communities can build upon in order to reduce the necessity of ad-hoc
expert opinion and increase transparency and reproducibility of harmonization
exercises. Furthermore, it provides an open-source, tested, and documented code
base which can be used and improved upon by these communities. Both of these are
clear steps in a positive direction for future climate and integrated assessment
modeling exercises.

### 2017-06-06 intro.tex
some final edits

Integrated Assessment Models (IAMs) are tools used to understand the complex
interactions between energy systems, economic systems, land use, the climate,
and air pollution, among other sectors . IAMs provide global projections of
systemic change by dividing the world into a number of representative regions
(typically 10 to 30), the definition of which is distinct for each model
\cite{krey_global_2014}. Results from IAMs are integral in a number of
international studies, which notably include projections of climate and economic
futures. Recently, flagship IAMs have developed scenarios based on the Shared
Socioeconomic Pathways (SSPs) \cite{van_vuuren_energy_2017, fricko_marker_2017,
  fujimori_ssp3:_2017, calvin_ssp4:_2017, kriegler_fossil-fueled_2017} which
quantify a variety of potential global futures. The SSPs are primarily designed
to being used in analyses of broad climate research which includes earth system
model (ESM) simulations, climate impact, adaptation climate mitigation
studies. They consequently underpin the United Nations' Intergovernmental Panel
on Climate Change (IPCC) across working groups.

While IAMs are implemented in myriad ways, including simulation and
optimization, the core inputs, outputs, and purpose are similar across different
models. Modeling teams incorporate existing data sets of energy systems, land
use, economic, demographic and emissions sources and concentrations, among other
data, in order to provide consistent historical trajectories of modeled
variables. The models then provide estimates of future trajectories of these
variables under various socioeconomic and technological assumptions as well as
proposed policy constraints, e.g., targets for future Greenhouse Gas (GHG)
emissions.

@@

With some frequency, however, historical data is updated to incorporate new
assumptions of source activity, new emissions accounting practices, or other
concerns. The global climate change community has recently developed a standard
historical emissions dataset for both anthropogenic emissions (i.e., the
Community Emissions Datasystem (CEDS)) \cite{hoesly_historical_2017} and
land-use change (LUC) emissions \cite{van_marle_historic_2017} as part of the
6th phase of the Coupled Model Inter-comparison Project (CMIP6). These
historical emissions trajectories in conjunction with future trajectories as
provided by IAMs are critical inputs to the climate-related modeling exercises
of CMIP6.

@@

In previous studies, e.g. the Representative Concentration Pathways (RCPs), IAM
results have been harmonized by sector and the 5 RCP global regions
\cite{vuuren_representative_2011}. Climate modeling efforts have continued to
progress, however, demanding increased spatial and sectoral resolution. In order
to address the growing dimensionality of model outputs while still providing a
consistent and scientifically rigorous harmonization procedure, an automated
process for determining harmonization methods is needed. The remainder of this
paper describes the methodology of the \code{aneris} code base
\cite{matthew_gidden_2017_802832}, written in the Python programming language
(detailed documentation available \href{http://mattgidden.com/aneris/}{online}),
which implements such a system in Section \ref{sec:meths}. The results of
applying the automated harmonization mechanism on two example IAM scenarios, one
with emissions growth and another with emissions mitigation, is presented in
Section \ref{sec:results}. Finally, the general effectiveness and potential
future improvements on the automated methodology is discussed in Section
\ref{sec:future}.

>>>>

Integrated Assessment Models (IAMs) are tools used to understand the complex
interactions between energy systems, economic systems, land use, the climate,
and air pollution, among other sectors . IAMs provide global projections of
systemic change by dividing the world into a number of representative regions
(typically 10 to 30), the definition of which is distinct for each model
\cite{krey_global_2014}. Results from IAMs are integral in a number of
international studies, which notably include projections of climate and economic
futures. Recently, flagship IAMs have developed scenarios based on the Shared
Socioeconomic Pathways (SSPs) \cite{van_vuuren_energy_2017, fricko_marker_2017,
  fujimori_ssp3:_2017, calvin_ssp4:_2017, kriegler_fossil-fueled_2017} which
quantify a variety of potential global futures. The SSPs are primarily designed
to be used in analyses of broad climate research which include earth system
model (ESM) simulations, climate impact, adaptation climate mitigation
studies. They consequently underpin the United Nations' Intergovernmental Panel
on Climate Change (IPCC) across working groups.

While IAMs are implemented in myriad ways, including simulation and
optimization, the core inputs, outputs, and purpose are similar across different
models. Modeling teams incorporate existing data sets of energy systems, land
use, economics, demographics and emissions sources and concentrations, among other
data, in order to provide consistent historical trajectories of modeled
variables. The models then provide estimates of future trajectories of these
variables under various socioeconomic and technological assumptions as well as
proposed policy constraints, e.g., targets for future Greenhouse Gas (GHG)
emissions.

@@

With some frequency, however, historical data is updated to incorporate new
assumptions of source activity, new emissions accounting practices, or other
concerns. The global climate change community has recently developed a standard
historical emissions dataset for both anthropogenic emissions (i.e., the
Community Emissions Datasystem (CEDS)) \cite{hoesly_historical_2017} and
land-use change (LUC) emissions \cite{van_marle_historic_2017} which, in
conjunction with the SSP IAM trajectories, will be used for climate-related
modeling exercises of CMIP6.

@@

In recent studies, i.e., the Representative Concentration Pathways (RCPs), IAM
results have been harmonized by sector and the 5 RCP global regions
\cite{vuuren_representative_2011}. Climate modeling efforts have continued to
progress, however, demanding increased spatial and sectoral resolution from
IAMs. In order to address the growing dimensionality of model outputs while
still providing a consistent and scientifically rigorous harmonization
procedure, an automated process for determining harmonization methods is
needed. The remainder of this paper describes the methodology of the
\code{aneris} code base \cite{matthew_gidden_2017_802832}, written in the Python
programming language (detailed documentation available
\href{http://mattgidden.com/aneris/}{online}), which implements such a system in
Section \ref{sec:meths}. The results of applying the automated harmonization
mechanism on two example IAM scenarios, one with emissions growth and another
with emissions mitigation, is presented in Section \ref{sec:results}. Finally,
the general effectiveness and potential future improvements on the automated
methodology is discussed in Section \ref{sec:future}.

### 2017-06-06 meth.tex
some final edits

IAM emission results are provided along a number of different dimensions,
including, temporal (normally half decade or decade), spatial (i.e., model
regions), gas species, and sectoral dimensions. Each individual temporal
trajectory, i.e., unique combinations of regions ($r$), species ($g$), and
sectors ($s$), must be harmonized to the initial time step. Given a model
trajectory, $m_{r, g, s}(t)$, historical trajectory, $h_{r, g, s}(t)$, and
initial time step $t_i$, a harmonized trajectory can be calculated. The
``goodness of fit'' of a trajectory depends on a number of factors, most notably
the faithful representation of the original, unharmonized, trajectory. However,
other important factors must be considered, such the representation of negative
trajectories (i.e., if a trajectory becomes negative, both the timing and total
magnitude should be as close as possible) which are of critical importance for
cumulative CO2 calculations.

In previous studies \cite{meinshausen_rcp_2011,rogelj_discrepancies_2011}, two
\textit{families} of methods have been used: those that operate on the ratio of
base year values (i.e., $\frac{h(t_i)}{m(t_i)}$) and those that operate on the
offset of base year values (i.e., $h(t_i) - m(t_i)$). The \code{aneris} code
base provides a number of variations on the classic methods, including
ratio-convergence methods shown in Equation \ref{eqs:ratio}, offset-convergence
methods shown in Equation \ref{eqs:offset}, and interpolation methods shown in
Equation \ref{eqs:interpolate}; the convergence factor, $\beta$, for associated
methods is shown first in Equation \ref{eqs:factor}. Note that model and
regional indices have been dropped for clarity. Each is a function of time,
model trajectory, historical trajectory, harmonization time step, and a
convergence time step ($t_f$), at which point the harmonized trajectory
converges to the unharmonized trajectory.

@@

The \code{aneris} code base provides a number of methods to choose from for each
of the harmonization families. A summary of all available methods is provided
below in Table \ref{tab:meths}.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.png}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
      library. For all decisions, upper branches represent a ``yes'' response
      and lower branches represent a ``no'' response. Decision-making thresholds
      are described below.}
  \end{center}
\end{figure}

@@

Variability of the historical trajectory is also an important characteristic
when considering the choice of harmonization method. Such trajectories are
normally indicative of emissions from land use sectors. Land use sectors are
modeled with varying fidelity by IAMs, but in general proxies are used for
emissions trajectories. Take for example emissions from grassland burning: for
most models, the use of grass and paturelands is endogenously modeled, and
emissions from the burning of these lands is determined by applying a emissions
factor; in general, the more grassland usage that occurs in a scenario, the
higher the emissions from the burning thereof. Therefore, consistency in
harmonization method is important because the effects are modeled similarly
across regions and species. In order to reliably detect land-use-like emissions,
a measure of the coefficient of variation of the first derivative of the
historical trajectory is calculated as shown in Equation \ref{eqs:cov}, where
$\sigma$ is the standard deviation, $\mu$ is the mean, and $c_v$ is the
coefficient of variation.

@@

Input data then undergoes a cleaning operation, which adds (null) model
trajectories that exist in the historical dataset but are not provided by the
model input and detects any issues that would cause the harmonization process to
fail. The methods used to harmonize the data are then determined and the
harmonization process is executed. Upon completion of the harmonization process,
aggregation of common analysis regions is performed (e.g., the 5 SRES
regions). Finally, any exogenous trajectories the user provides are
added. Exogenous trajectories are normally provided for unmodeled gases with
well-accepted scenario trajectories, e.g., chlorofluorocarbons provided by WMO
\cite{wmo2014}. Upon completion, the harmonized trajectories and meta data
regarding the harmonization process are returned. A description of all returned
meta data is provided in Table \ref{tab:metadata}.

>>>>

IAM emission results are provided along temporal (normally half decade or
decade), spatial (i.e., model regions), gas species, and sectoral
dimensions. Each individual temporal trajectory, i.e., unique combinations of
regions ($r$), species ($g$), and sectors ($s$), must be harmonized to the
initial time step. Given a model trajectory, $m_{r, g, s}(t)$, historical
trajectory, $h_{r, g, s}(t)$, and initial time step $t_i$, a harmonized
trajectory can be calculated. The ``goodness of fit'' of a trajectory depends on
a number of factors, most notably the faithful representation of the original,
unharmonized, trajectory. However, other important factors must be considered,
such as the representation of negative trajectories (i.e., if a trajectory
becomes negative, both the timing and total magnitude should be as close as
possible) which are of critical importance for cumulative CO2 calculations.

In previous studies \cite{meinshausen_rcp_2011,rogelj_discrepancies_2011}, two
\textit{families} of methods have been used: those that operate on the ratio of
base year values (i.e., $\frac{h(t_i)}{m(t_i)}$) and those that operate on the
offset of base year values (i.e., $h(t_i) - m(t_i)$). The \code{aneris} code
base provides a number of variations on the classic methods, including
ratio-convergence methods shown in Equation \ref{eqs:ratio}, offset-convergence
methods shown in Equation \ref{eqs:offset}, and interpolation methods shown in
Equation \ref{eqs:interpolate}; the convergence factor, $\beta$, for associated
methods is shown first in Equation \ref{eqs:factor}. Note that region, species,
and sector indices have been dropped for clarity. Each equation is a function of
time, model trajectory, historical trajectory, harmonization time step, and a
convergence time step ($t_f$), at which point the harmonized trajectory
converges to the unharmonized trajectory. The \code{aneris} code base provides a
number of methods to choose from for each of the harmonization families. A
summary of all available methods is provided in Table \ref{tab:meths}.

@@

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.png}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
      library. For all decisions, upper branches represent a ``yes'' response
      and lower branches represent a ``no'' response. $dH$ is defined as
      $\left|\frac{h(t_i) - m(t_i)}{h(t_i)}\right|$, and decision-making
      thresholds are described below. Finally, default choices in \textit{green}
      label methods that will likely closely match unharmonized results, choices
      in \textit{yellow} will likely somewhat match unharmonized results, and
      choices in \textit{red} can be expected to have a large relative
      difference between harmonized and unharmonized results.}
  \end{center}
\end{figure}

@@

Variability of the historical trajectory is also an important characteristic
when considering the choice of harmonization method. Such trajectories are
normally indicative of emissions from land use sectors. Land use sectors are
modeled with varying fidelity by IAMs, but in general proxies are used for
emissions trajectories. Take for example emissions from grassland burning: for
most models, the use of grass and paturelands is endogenously modeled, and
emissions from the burning of these lands is determined by applying an emissions
factor; in general, the more grassland usage that occurs in a scenario, the
higher the emissions from the burning thereof. Therefore, consistency in
harmonization method is important because the effects are modeled similarly
across regions and species. In order to reliably detect land-use-like emissions,
a measure of the coefficient of variation of the first derivative of the
historical trajectory, $c_v$, is calculated as shown in Equation \ref{eqs:cov},
where $\sigma$ is the standard deviation, and $\mu$ is the mean of the
trajectory.

@@

Input data then undergoes a cleaning operation, which adds (null) model
trajectories that exist in the historical dataset but are not provided by the
model input and detects any issues that would cause the harmonization process to
fail. The methods used to harmonize the data are then determined and the
harmonization process is executed. Upon completion of the harmonization process,
aggregation of common analysis regions is performed (e.g., the 5 RCP
regions). Finally, any exogenous trajectories the user provides are
added. Exogenous trajectories are normally provided for unmodeled gases with
well-accepted scenario trajectories, e.g., chlorofluorocarbons provided by WMO
\cite{wmo2014}. Upon completion, the harmonized trajectories and meta data
regarding the harmonization process are returned. A description of all returned
meta data is provided in Table \ref{tab:metadata}.

### 2017-06-06 results.tex
some final edits

In order to test the \code{aneris} harmonization procedure, results from the IAM
MESSAGE-GLOBIOM \cite{fricko_marker_2017} was processed. Two scenarios from the
SSP scenario library \cite{Riahi2017153,Rao2017346} are investigated. The SSP2,
or ``middle of the road'', scenario family is chosen to be analyzed because
MESSAGE-GLOBIOM is the marker scenario for this SSP. Of all SSP2 scenarios, the
reference (marker) and SSP2-45 scenarios are analyzed. Notably, emissions from
multiple sectors tend to increase in the reference scenario, thus testing the
harmonization mechanism with an increasing emissions pathway. The SSP-45
scenario, where the ``45'' designation refers to a scenario with end-of-century
emissions resulting in approximately a 4.5 $\frac{\text{W}}{\text{m}^2}$
radiative forcing due to both warming from GHGs and fluoridated gases as well as
cooling from aerosols, is chosen because mitigation technologies and policies
are enacted causing a general reduction in pollutants and GHGs, including
(eventual) negative CO2 emissions in some regions and sectors due to carbon
capture and sequestration and afforestation. This scenario explores the
harmonization mechanism's response to generally decreasing emission
trajectories.

MESSAGE-GLOBIOM includes a representation of 11 distinct regions which can be
mapped directly to the 5-region definition used in the SRES. Historical data is
taken from previously described LUC and anthropogenic sources, which comprise 10
separate pollutant and GHG species and 13 sectors. Therefore, a total of 1430
distinct trajectories were harmonized for each scenario.

>>>>

In order to test the \code{aneris} harmonization procedure, results from the IAM
MESSAGE-GLOBIOM \cite{fricko_marker_2017} was processed. Two scenarios from the
SSP scenario library \cite{Riahi2017153,Rao2017346} are investigated. The SSP2,
or ``middle of the road'', scenario family is chosen to be analyzed because
MESSAGE-GLOBIOM is the marker scenario for this SSP. Of all SSP2 scenarios, the
reference (marker) and SSP2-45 scenarios are analyzed. Notably, emissions from
multiple sectors tend to increase in the reference scenario, thus testing the
harmonization mechanism with an increasing emissions pathway. The SSP2-45
scenario, where the ``45'' designation refers to a scenario with end-of-century
emissions resulting in approximately a 4.5 $\frac{\text{W}}{\text{m}^2}$
radiative forcing due to both warming from GHGs and fluoridated gases as well as
cooling from aerosols, is chosen because mitigation technologies and policies
are enacted causing a general reduction in pollutants and GHGs, including
(eventual) negative CO2 emissions in some regions and sectors due to carbon
capture and sequestration and afforestation. This scenario explores the
harmonization mechanism's response to generally decreasing emission
trajectories.

MESSAGE-GLOBIOM includes a representation of 11 distinct regions which can be
mapped directly to the 5-region definition used in the RCPs. Historical data is
taken from previously described LUC and anthropogenic sources, which comprise 10
separate pollutant and GHG species and 13 sectors. Therefore, a total of 1430
distinct trajectories were harmonized for each scenario.
